Linear stability of magnetized massive protoplanetary disks

### 2014-04-03 paper/paper.tex
tex updates

\begin{abstract}
  Magneto-rotational instability (MRI) and gravitational instability
  (GI) are the two principle routes to turbulent angular momentum
  transport in accretion disks. Protoplanetary disks may develop 
  both. This paper aims to reinvigorate interest in the study of
  magnetized  massive protoplanetary disks, starting from the basic
  issue of stability. The local linear stability of a 
  self-gravitating, uniformly magnetized, differentially rotating, 
  three-dimensional disk subject to axisymmetric perturbations is
  calculated numerically. The formulation includes 
  resistivity. It is found that the reduction in the disk scale height
  by self-gravity (SG) can decrease MRI growth rates. SG causes the MRI to
  become global in the vertical direction, and MRI modes
  with small radial lengthscales are stabilized. In massive
  disks with layered resistivity, the MRI is not well-localized
  to regions where the Elsasser number exceeds unity. For MRI modes
  with radial lengthscales on the order of the disk scale-height, SG
  can enhance density perturbations, an effect that becomes 
  significant in the presence of a strong toroidal field. In
  gravitationally unstable disks where GI and MRI growth rates are
  comparable, the character of unstable modes can transition smoothly
  between MRI and GI. Implications for non-linear simulations are
  discussed briefly.

>>>>

\begin{abstract}
  Magneto-rotational instability (MRI) and gravitational instability
  (GI) are the two principle routes to turbulent angular momentum
  transport in accretion disks. Protoplanetary disks may develop 
  both. This paper aims to reinvigorate interest in the study of
  magnetized  massive protoplanetary disks, starting from the basic
  issue of stability. The local linear stability of a 
  self-gravitating, uniformly magnetized, differentially rotating, 
  three-dimensional disk subject to axisymmetric perturbations is
  calculated numerically. The formulation includes 
  resistivity. It is found that the reduction in the disk thickness
  by self-gravity (SG) can decrease MRI growth rates; the MRI
  becomes global in the vertical direction, and MRI modes 
  with small radial lengthscales are stabilized. In massive
  disks with layered resistivity, the MRI is not well-localized
  to regions where the Elsasser number exceeds unity. For MRI modes
  with radial lengthscales on the order of the disk thickness, SG
  can enhance density perturbations, an effect that becomes 
  significant in the presence of a strong toroidal field. In
  gravitationally unstable disks where GI and MRI growth rates are
  comparable, the character of unstable modes can transition smoothly
  between MRI and GI. Implications for non-linear simulations are
  discussed briefly.

### 2014-04-03 paper/results_azi.tex
tex updates

\section{Effect of an azimuthal field}\label{result3}
In this section we add an azimuthal field so that $B_y\neq 0$,  
parameterized by $\epsilon \equiv B_y/B_z$. However, we continute to
use $B_z$ for normalizations and $\beta$ is associated with the
vertical Alfven speed. We also extend the previous calculations to the
full disk $z\in[-Z_s,Z_s]$, which allows us to compare the effect of
self-gravity on MRI modes with different symmetries across the
midplane.

\subsection{Ideal disks with MRI} 
Here we consider an isothermal disk with $Q=0.2$
($Q_\mathrm{2D}=0.72$) and $\beta=10$ in the limit of ideal MHD
($\Lambda_0=100$). Gravitational instability is not expected because 
Fig. \ref{compare_growth3} shows that even for $Q=0.18$, GI is 
surpressed for $\beta \lesssim 15$. We consider azimuthal field
strengths with $\epsilon\in[0,3]$.

Fig. shows the MRI growth rates for the above system as a function of
$k_x$. The top panel are modes with even midplane symmetry: 
$W^\prime(0)= 0$ for $\epsilon\to0$, and the bottom panel are modes with odd midplane 
symmetry: $W(0) = 0$ for $\epsilon\to0$. The latter set of modes were 
excluded in the previous sections by midplane boundary 
conditions. The plot also include growth rates computed in the Cowling
approximation. As expected,  $\avg{E_g}<\avg{E_m}$, so none of the
modes are energetically dominated by self-gravity.

%no GI for odd modes

%general stabilization of MRI

@@

Consider first the even modes in the top panel of 
Fig. \ref{compare_growth3_tilted}. %Self-gravity slightly stabilizes
%modes with small $k_x$ as $B_y$ is increased. 
Self-gravity destabilizes modes with  $k_xH\gtrsim O(1)$. Consequently, the
cut-off wavenumber is larger when SG is included. %get smaller scale
                                %MRI 
Destablization is most effective for purely vertical fields: with
$\epsilon=0,\, 
k_xH\simeq 1.4$, SG increases the growth rate by $\sim 30\%$. Although
these modes are fundamentally magnetic, this is consistent with
\cite{goldreich65a}, who find SG can only destabilize 
symmetric density perturbations, which is the case for $B_y=0$ here.  
With increasing $\epsilon$, we find the density perturbation $W$ deviates from an even 
function, $|W(0)|$ decreases and $\max{(|W|)}$ moves off of the mid-plane.  
%density pert at the midplane goes down
%We expect this to reduce the destabilization
%effect of SG,  
Together with the increased total magnetic pressure with 
$\epsilon$, destabilization by SG weakens. 
Thus, the Cowling approximation becomes increasingly good with stronger
$B_y$ for the even modes.

@@

The modes in the bottom panel of Fig. \ref{compare_growth3_tilted} 
display opposite behavior. For $B_y=0$, $W$ is odd and 
self-gravity has no effect. When $B_y>0$, we find $W$ deviates from an odd
function and the mid-plane density perturbation $|W(0)|$ increases. 
SG is stabilizing for these modes at all wavelengths, and is most
effective at $k_xH =  O(1)$.  
%destabilizing effect is absent  
%for $\epsilon=0,\, k_xH\simeq 1.2$, SG decreases the growth rate by $\sim
%13\%$.

Fig. \ref{result_tilted} show eigenfunctions for $\epsilon=3$ and
$k_xH=1.2$ with and without the Cowling approximation. These differ
from the even modes considered above in that the horizontal magnetic
perturbations maximize at $z=0$ (cf. Fig. \ref{mri_massive_resis}).  
SG significantly enhances the mid-plane density perturbation, making the
gravitational potential energy comparable to the magnetic energy, 
which becomes more confined near the mid-plane.

@@

%strong field, compressibility 
To interpret the above result for odd modes, note 
that compressibility affects the MRI in the presence of an 
azimuthal field, even in a non-self-gravitating disk. If the perturbed 
disk remains in vertical hydrostatic equilibrium, then 
\begin{align} 
  |W|\sim \frac{B_y}{\mu_0\rho}|\dby|,
\end{align}
to order of magnitude in a non-SG disk, so a strong toroidal field can
cause a large density perturbation
\citep{pessah05}. Fig. \ref{result_tilted} shows that this is
amplified by self-gravity.

We checked that for the modes in Fig. \ref{result_tilted} the vertical
velocity is small, $|\dvz|/\left(|\dvx|^2+|\dvy|^2\right)^{1/2}
\lesssim 0.2$, so the above argument applies. Compressibilty is 
enhanced by both the azimuthal field and self-gravity, which tends
to stabilize the MRI \citep{kim00}. Its effect is important here  
because the azimuthal Alfven speed is sonic. The destabilization
effect of SG is absent (or unimportant) because the density
perturbation is not symmetric.

@@

\subsection{Resistive disks with GI}
Here we examine an isothermal, resistive disk with $Q=0.18$ and
$\Lambda_0=0.1$. GI is permitted, and has a similar growth rate as the
MRI.

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 2cm 0cm
    0cm]{figures/compare_growth3_tilted_resis_ep0.ps}  
  \includegraphics[width=\linewidth,clip=true,trim=0cm 2cm 0cm
    0.52cm]{figures/compare_growth3_tilted_resis_ep1.ps}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
    0.52cm]{figures/compare_growth3_tilted_resis_ep2.ps}
  \caption{Growth rates in isothermal resistive disks with $Q=0.18$
    ($Q_\mathrm{2D}=0.67$) and $\Lambda_0=0.1$.  
    \label{compare_growth3_tilted_resis}}
\end{figure}

>>>>

\section{Effect of an azimuthal field}\label{result3}
In this section we use the setup of case 2 described in
\S\ref{linear}, and examine the effect of an azimuthal field so that
$B_y\neq 0$,   
parameterized by $\epsilon \equiv B_y/B_z$. However, we continute to
use $B_z$ for normalizations and $\beta$ is associated with the
vertical Alfven speed. We also extend the previous calculations to the
full disk $z\in[-Z_s,Z_s]$,  which allows us to compare the effect of
self-gravity on MRI modes with different symmetries across the
midplane.

\subsection{Ideal disks with MRI} 
We consider an isothermal disk with $Q=0.2$ ($Q_\mathrm{2D}=~0.72$) and
$\beta=10$ in the limit of ideal MHD ($\Lambda_0=100$). Gravitational
instability is not expected because Fig. \ref{compare_growth3} shows
that even for $Q=0.18$, GI is surpressed for $\beta \lesssim 15$. %We
%consider azimuthal field strengths with $\epsilon\in[0,3]$.

Fig. \ref{compare_growth3_tilted} show MRI growth rates for
$B_y/B_z=0,\,1,\,2$ and $3$. We divide the modes into two catagories
depending on the extremum of magnetic energy at the midplane. 
The top panel are modes where $E_m$ has a local minimum at $z=0$ and 
the bottom panel are modes where $E_m$ has a local maximum at $z=0$. 
The latter set of modes were excluded in the previous sections by
midplane boundary conditions. 
%The top panel are modes with even midplane symmetry: 
%$W^\prime(0)= 0$ for $\epsilon\to0$, and the bottom panel are modes with odd midplane 
%symmetry: $W(0) = 0$ for $\epsilon\to0$. 
We also plot growth rates computed in the Cowling
approximation. As expected,  $\avg{E_g}<\avg{E_m}$, so none of the
modes are energetically dominated by self-gravity.

%no GI for odd modes 
%general stabilization of MRI

@@

Consider first modes in the top panel of 
Fig. \ref{compare_growth3_tilted}. As with previous results,  
self-gravity destabilizes modes with  $k_xH\gtrsim O(1)$. Consequently, the
cut-off wavenumber is larger when SG is included. %get smaller scale
                                %MRI 
Destablization is most effective for purely vertical fields: with
$\epsilon=0,\, k_xH\simeq 1.4$, SG increases the growth rate by $\sim
30\%$. For $B_y=0$ we find the density perturbation $W(z)$ is an even
function. Although these modes are fundamentally magnetic, this is consistent with
\cite{goldreich65a}, who showed that SG can only destabilize 
symmetric density perturbations.  
With increasing $\epsilon$, we find $W$ deviates from an even
function. %,% $|W(0)|$ decreases and $\max{(|W|)}$ 
%moves off of the mid-plane.    
%density pert at the midplane goes down
%We expect this to reduce the destabilization
%effect of SG,  
Together with the increased total magnetic pressure with 
$\epsilon$, destabilization by SG weakens. 
Thus, the Cowling approximation becomes increasingly good with stronger
$B_y$ for these modes.

@@

The modes in the bottom panel of Fig. \ref{compare_growth3_tilted} 
display opposite behavior. For $B_y=0$ we find $W$ is odd, and
self-gravity has no effect. When $B_y>0$, $W$ deviates from an
odd function and the mid-plane density perturbation $|W(0)|$
increases. SG is stabilizing for these modes at all wavelengths, and
is most effective at $k_xH =  O(1)$.  
%destabilizing effect is absent  
%for $\epsilon=0,\, k_xH\simeq 1.2$, SG decreases the growth rate by $\sim
%13\%$.

Fig. \ref{result_tilted} show eigenfunctions for $\epsilon=3$ and
$k_xH=1.1$ with and without the Cowling approximation. 
%These differ  
%from the even modes considered above in that the magnetic energy
%maximize at $z=0$ (cf. Fig. \ref{mri_massive_resis}).  
SG significantly enhances the midplane density perturbation, making the
gravitational potential energy comparable to the magnetic energy, 
which becomes more confined near the mid-plane.

@@

%strong field, compressibility 
To interpret the above result for modes with magnetic energy
concentrated at the midplane, we note that compressibility affects the
MRI in the presence of an azimuthal field even in a
non-self-gravitating disk. If the perturbed disk remains in vertical
hydrostatic equilibrium, then  
\begin{align} 
  |W|\sim \frac{B_y}{\mu_0\rho}|\dby|,
\end{align}
to order of magnitude in a non-SG disk. Thus a strong azimuthal field can 
cause a large density perturbation \citep{pessah05}. We checked that
for the modes in Fig. \ref{result_tilted}, vertical velocities are 
small, $|\dvz|/\left(|\dvx|^2+|\dvy|^2\right)^{1/2} \lesssim 0.2$.

Compressibilty is enhanced by an azimuthal field, which has a
stabilizing effect on the MRI \citep{kim00}. %because...?
This effect is significant for $\epsilon=3$ because the azimuthal
Alfven speed is sonic.    
%maybe blaes 
Fig. \ref{result_tilted} indicates that self-gravity further enhances
compressibility, and therefore stabilization. We suspect this is
overwhelmed by the destabilization effect of SG because the density
perturbation has an anit-symmetric component.

%compressibility -> pressure restoring force

@@

%\subsection{Ideal disks with GI (TBC)}

\subsection{Resistive disks with GI}
Here we consider an isothermal, resistive disk with $Q=0.18$,
$\Lambda_0=0.1$ and $\beta=100$. This system permits GI and MRI. 
%GI is permitted, and has a similar
%growth rate as the  
%MRI. 
Fig. \ref{compare_growth3_tilted_resis} show growth rates for
$\epsilon=0,\,1$ and $2$. For $B_y=0$, MRI and GI are decoupled except 
for a narrow range of $k_x$ in which the lower MRI modes transitions to
GI. Notice that the upper MRI modes intersect the GI branch. There is no
interaction because the upper MRI modes have anti-symmetric $W(z)$ whereas
the GI modes have symmetric $W(z)$.

%GI, and there are MRI modes available on that scale

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 2cm 0cm
    0cm]{figures/compare_growth3_tilted_resis_ep0.ps}  
  \includegraphics[width=\linewidth,clip=true,trim=0cm 2cm 0cm
    0.52cm]{figures/compare_growth3_tilted_resis_ep1.ps}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
    0.52cm]{figures/compare_growth3_tilted_resis_ep2.ps}
  \caption{Growth rates in isothermal resistive disks with $Q=0.18$
    ($Q_\mathrm{2D}=0.67$), $\beta=100$ and $\Lambda_0=0.1$. For
    $B_y=0$, the upper MRI modes have anti-symmetric density
    perturbations, $W(0) = 0$; while the lower MRI modes have symmetric
    density perturbations, $W^\prime(0) = 0$. For $B_y/B_z=2$ the
    overstable modes have non-zero real frequencies. 
    \label{compare_growth3_tilted_resis}}
\end{figure}

Introducing $B_y = B_z$ leads to an exchange in the mode
characters. For $k_xH\lesssim 0.9$ the modes on the two MRI branches
are similar to the vertical field case. However, for $k_xH\gtrsim0.9$ 
the upper MRI mode transitions to GI, for which $E_m(0)$ is a
minimum; and the lower MRI mode has $E_m(0)$ being a maximum. 
%interchange possible because of GI branch

%In both cases where the gravitational energy is significant, the
%density perturbation is symmetric.

Increasing the azimuthal field further to $B_y=2B_z$ we find
overstable MRI modes \citep{gammie96} with non-negligible real
frequencies. An example is shown in
Fig. \ref{result_tilted_overstable}. Notice the density/potential
perturbation is off-set from the midplane. This is not possible for
pure GI \citep{goldreich65a}. Thus, these overstable MRI modes indeed
become self-gravitating, before being stabilized. 
%these may lead to significant vertical motions

Notice in Fig. \ref{compare_growth3_tilted_resis} the disappearence of
magnetic modes as $B_y$ is increased. For $B_y=2B_z$, MRI and GI
operate at distinct scales. This implies that regions unstable to GI
cannot develop MRI.

%This may lead to 
%the lower MRI mode transitions to the top MRI mode

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
    0cm]{figures/result_tilted_resis.ps}
  \caption{Overstable MRI mode in an isothermal resistive disk with
    $Q=0.18 (Q_\mathrm{2D} = 0.67)$, $\Lambda_0=0.1$ and $\beta=100$. 
    The mode has a real frequency $\omega = 0.059\Omega$, or
    $\omega/\gamma \simeq 0.2 $.   
    \label{result_tilted_overstable}}
\end{figure}

### 2014-04-03 paper/results_gi.tex
tex updates

\subsection{Co-existence of MRI and GI} 
Fig. \ref{compare_growth3} show growth rates for modes with $k_xH=1$
as a function of $\beta$ in disks with $Q=0.18$, $Q=0.14$ and
$Q=0.12$.   
All three cases display distinct GI modes (red/brown branch). The GI 
growth rates are $\gamma\simeq 0.25\Omega,\,0.6\Omega,\,0.8\Omega$ for
$Q=0.18,\,0.14,\,0.12$, respecitvely. GI is stabilized by magnetic
pressure for sufficiently small $\beta$, and the critical
field strength increases with increasing self-gravity \citep{nakamura83}. 
For $Q=0.18$, GI is stabilized for $\beta \lesssim 15$. Nevertheless, 
the MRI branch for $\beta < 15$ becomes self-gravitating, so that
density perturbations will still develop, even though GI does not
formally operate.

>>>>

\subsection{Co-existence of MRI and GI} 
Fig. \ref{compare_growth3} show growth rates for modes with $k_xH=1$
as a function of $\beta$ in disks with $Q=0.18$, $Q=0.14$ and
$Q=0.12$.   
All three cases display distinct GI modes (red/brown branch). The GI 
growth rates are $\gamma\simeq 0.25\Omega,\,0.6\Omega,\,0.8\Omega$ for
$Q=0.18,\,0.14,\,0.12$, respecitvely. GI is stabilized by magnetic
pressure for sufficiently small $\beta$. The critical
field strength increases with increasing self-gravity \citep{nakamura83}. 
For $Q=0.18$, GI is stabilized for $\beta \lesssim 15$. Nevertheless, 
the MRI branch for $\beta < 15$ becomes self-gravitating, so that
density perturbations will still develop, even though GI does not
formally operate.

### 2014-04-03 paper/summary.tex
tex updates

%SG through 
We find MRI perturbations with radial scale of $\sim H$ can acquire  
density perturbations in a massive but Toomre-stable disk. 
This occurs when the MRI is weak, for example with a strong field or
high resisitivty. We argue in that case $\lambda\simH$, so the density
perturbation caused by the MRI can be enhanced by self-gravity.

%further enhancing density perturbations. 
At this point it is worth mentioning previous non-linear calculations 
performed by \cite{kim03} and \cite{fromang04a,fromang04} in the
context of galactic and circumstellar disks, respectively. These
authors simulated the MRI in self-gravitating disks. They find
self-gravity has little impact on the MRI.

%cite{kim03} and \cite{fromang04}

%Numerical experiments
%varying the field strength and resisitivity indicate this is again a
%result

%GI/MRI: avoided crossings

@@

%\subsection{Implications for non-linear simulations}
%what scaes? resolution
%why don't formang  and kim/ostriker see any difference?

\subsection{Caveats and outlooks}
We discuss below two major extensions to our linear model that
should be undertaken, before performing non-linear simulations of
magnetized, self-gravitating disks, which is our eventual goal.

@@

%spiral arms, GI angular mometum distribution 
\emph{Beyond axisymmetry.} It is well known that self-gravitating
disks can be unstable to non-axisymmetric perturbations but stable
against axisymmetric perturbations, although the threshold Toomre
parameter for both is still of order unity. %papaloizou/lin,
                                %sajvoini??

>>>>

%SG through 
We find MRI perturbations with radial scale of $\sim H$ can acquire  
density perturbations in a massive but Toomre-stable disk. 
This occurs when the MRI is weak, for example with a strong field or
high resisitivty. We argue in that case $\lambda\sim H$, so the density
perturbation caused by the MRI can be enhanced by self-gravity.

%further enhancing density perturbations. 
At this point it is worth mentioning previous non-linear simulations
of magnetized self-gravitating galactic and circumstellar disks
\citep{kim03,fromang04a,fromang04}. These authors find self-gravity
does not enhance MRI density flucations significantly. However, they
employed ideal MHD simulations with gas-to-magnetic pressure ratios of
order $10^2$ to $10^3$. According  
to our results, self-gravity is not expected to affect the MRI in this
regime of $\beta$. On the other hand, \cite{fromang04} found
self-gravity makes MRI turbulence more coherent. This may be related
to our finding that small radial scale MRI is surpressed when
self-gravity is included in the background equilibria.

Physically, we expect MRI to only interact with self-gravity when
their spatial scales are similar. Because self-gravity acts globally in the
vertical direction, for it to affect the MRI, future non-linear
simluations should consider parameter regimes in which the MRI is
vertically global. Indeed, in the setup of \cite{kim03}, the disk
scale height exceeds the MRI vertical wavelength and self-gravity has
little effect.

%We suggest that in
%order for self-gravity to be significant,
%We suggest that in
%order for self-gravity to influence MRI,      
%cite{kim03} and \cite{fromang04}
%Numerical experiments
%varying the field strength and resisitivity indicate this is again a
%result

%GI/MRI: avoided crossings, relevance to simulations
Curiously, when GI and MRI are simultaneously supported, we find
unstable modes transition between MRI and GI. Such modes
have comparable potential and magnetic energy perturbations. These
transitions occur smoother with decreasing $\beta$
(Fig. \ref{compare_growth3}) or increasing $k_x$
(Fig. \ref{compare_growth3_Q01d2}). The latter implies that, in order
to capture magnetic-gravitational interactions represented by these
intermediate modes, non-linear simulations must resolve radial scales
smaller than the most unstable GI mode. For example, 
Fig. \ref{compare_growth3_Q01d2} suggest radial scales down to $\sim H/2$
should be resolved.

%This is
%consistent with the MRI only affected by self-gravity for strong
%fields 
%mode classificication ambgious 
%This is consistent /expected
%numerical simulations should measure a continous range of modes
%This is
%consistent self-gravity only affecting the MRI when it is weak.  
%mag field only affect GI through pressure
%mri influenced by GI at large kx - > small scale -> need resolution

@@

We examined the effect of an additional azimuthal field, while
keeping the vertical field at fixed strength. In this case, we also
relaxed the equitorial symmetry condition applied previously and
considered the full disk column. Self-gravity affects the MRI
differently depending on the midplane symmetry of the MRI density
perturbation. Self-gravity destabilizes MRI modes where the magnetic
energy reaches a minimum at $z=0$, which have a symmetric
density perturbation in the limit $B_y\to0$. However, 
self-gravity stabilizes MRI modes where the magnetic energy maximizes
at $z=0$. These modes have an anti-symmetric density
perturbation in the limit $B_y\to0$. This stabilization effect is
stronger for increasing $B_y$. Previous linear calculations indicate
increased compressibility associated with a toroidal field stabilizes
the MRI \citep{kim03}. We conjecture that self-gravity can
further enhance this effect.   
%can't do WKB (can't tell symmetries)

%lesure 14 (strong azimuthal field)
%implications non-linear MRI simulations with a strong toroidal field
%would over-estimate growth rates (and perhaps turbulence strengths)
%if SG is not accounted for. not applicable to self-gravitating disks     
%previous linear calcs show compressibility stabilizes MRI 
%additional pressure stabilizing MRI

\subsection{Caveats and outlooks}
We discuss below two major extensions to our linear model that
should be undertaken, before embarking on non-linear simulations of
magnetized, self-gravitating disks, which is our eventual goal.

@@

%spiral arms, GI angular mometum distribution 
\emph{Beyond axisymmetry.} It is well known that self-gravitating
disks can be unstable to non-axisymmetric perturbations but stable
against axisymmetric perturbations \citep{papaloizou89,papaloizou91}.
